Semi-supervised low-rank logistic regression for high-dimensional neuroimaging data

### 2015-05-17 paper_nips/paper.tex
solver passage

\documentclass{article} % For LaTeX2e
\usepackage{nips14submit_e,times}
\usepackage{hyperref}
\usepackage{url}
\usepackage{amsmath,amsfonts,amsthm}
\usepackage{bbm}
\usepackage{algorithm,algorithmic}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{bbm}
\usepackage[titletoc]{appendix}

@@

\title{Semi-supervised low-rank logistic regression for
high-dimensional neuroimaging datasets}

@@

\author{Danilo Bzdok, Michael Eickenberg, Bertrand Thirion, Ga\"el
  Varoquaux \\\\\textbf{\textit{email:} }firstname.lastname@inria.fr}

@@

\section{Introduction}
%
Methods used in neuroimaging research can be grouped by discovering
neurobiological structure or revealing the neural correlates associated
with mental tasks.
To discover coherent distributions of activation structure across time,
independent component analysis (ICA; \cite{beckmann2005}) is often used
to decompose the BOLD (blood-oxygen level-dependent) signals into the
important modes of variation.
The ensuing spatial activation patterns are believed to represent
brain networks of
functionally interacting brain regions.
Similarly, sparse PCA (SPCA; \cite{varoqu2011}) has been used to
separate brain activity signals into parsimonious network components.
Thus extracted brain networks have been shown to be
manifestations of electrophysiological oscillation frequencies \cite{hipp15}.
Their fundamental role is
attested by continued covariation during sleep and anesthesia.
%
Network discovery is typically performed by applying ICA or SPCA on
unlabeled "resting-state" data. These capture brain dynamics
during ongoing random thought without controlled environmental input.
The rationale is that the biggest fraction of BOLD signals do
not correlate with a particular behavior, stimulus, or experimental task.

@@

On the other hand, to investigate
the neural correlates underlying mental tasks,
the general linear model (GLM; \cite{friston94}) is the dominant approach.
The contribution of
individual brain voxels is estimated
according to a design matrix of experimental tasks.
Alternatively, psychophysiological interactions (PPI; \cite{friston97}),
elucidate the functional interactions between voxels as a function
of experimental tasks.
Dynamic causal modeling (DCM; \cite{stephan04}), in turn, quantifies directed,
task-driven influences between regions
by treating the brain as a nonlinear dynamic system with unknown
ni  neuronal states. As a last example, always more neuroimaging studies model
experimental tasks by training classification algorithms on brain signals
\cite{poldrack09decoding}.
All these methods are applied to labeled task data that capture brain dynamics
during stimulus-guided behavior.
Two important conclusions can be drawn.
First, the mentioned supervised neuroimaging analyses operate
without exception in a single-voxel space. This ignores the fact that the BOLD
signal contains coherent spatial activation patterns.
Second, existing neuroimaging analyses do not acknowledge the fact that the
task-induced changes of the BOLD signal amount to less than 5\%
of baseline activity \cite{fox07}. They do thus not exploit the high similarity
of BOLD dynamics of the human brain at rest and during experimental tasks.
Indeed, similar brain networks were observed when applying ICA on
large quantities of
rest and task data \cite{smith2009}.

@@

Both aspects can be conjointly exploited in a 
semi-supervised (i.e., use rest and task data)
low-rank (i.e., perform network decomposition)
approach to neuroimaging analysis.
%
The integration of brain-network discovery in a 
supervised classification goal should identify the
neurobiological structure that allows for the best predictive models.
%
Autoencoders suggest themselves because they can emulate
variants of most nonsupervised learning algorithms,
including PCA, SPCA, and ICA.
Autoencoders
are one-layered learning models that find local and global
representation from the inputs
by improving reconstructions from them \cite{hinton06}.
%
They behave like a PCA
in case of one linear hidden layer and a squared error loss
\cite{baldi1989neural}.
This architecture yields a convex optimization objective
with unique global minimum.
Autoencoders behave like a SPCA if shrinkage terms are added for the
matrix weights in the optimization objective.
In turn, they behave like an ICA in case of a nonlinear convex
function of the first-layer activation and tied weights \cite{le2011ica}.
These authors further demonstrated that ICA, sparse autoencoders, and 
sparse coding are mathematically equivalent
under mild conditions.
In this way, autoencoders can flexibly project the neuroimaging data
onto the axes of main variation and thus
reverse-engineer properties of the data-generating
neural process \cite{olshausen96}.

@@

In the present investigation,
an autoencoder will be fed by (unlabeled) rest data and
integrated as bottleneck
into a low-rank logistic regression fed by (labeled) task data.
Using the chain rule in back-propagation, we can then
solve the unsupervised data representation and a supervised classification
in an identical statistical learning problem.
%
From the perspective of dictionary learning, the first layer can be
viewed as a learned set of basis functions
whose linear combinations are learned
in the second layer \cite{olshausen96}.
%
Neurobiologically, this allows 
delineating a low-dimensional manifold of brain network patterns and then 
distinguishing mental tasks
by their most discriminative linear combinations.
%
Theoretically, a big reduction of the model variance is expected by using the
resting-state autoencoder as regularization mechanisms
to put probability mass on the most neurobiological
valid members of the model space.
%
The generalization performance should consequently be improved due to 
much reduced Vapnik-Chervonenkis dimensions of the classification estimator.
%
Taken together,
the important modes of variation in brain dynamics and
the neural correlates subserving mental operations
have mostly been studied in isolation.
We provide a principled computational framework to link these previously
unconnected domains of systems neuroscience.

@@

linear decomposition + elastic net penalty -> has characteristics of a
sparse PCA

@@

AE:
coordinate system for points on the manifold

@@

LR:

@@

@@

RMSprop:
batch size:

@@

theano/python

\section{Experimental Results}
The source code, which can reproduce the most important results and
figures, is available at \url{http://github.com/banilo/nips2015}.

@@

@@

\section{Discussion and Conclusion}
%

@@

There is much uncertainty about the most pertinent representation
of neural activation information

%
\paragraph{Acknowledgment}
{\small
Data were provided the Human Connectome Project.
}

>>>>

\documentclass{article} % For LaTeX2e
\usepackage{nips14submit_e,times}
\usepackage{hyperref}
\usepackage{url}
\usepackage{amsmath,amsfonts,amsthm}
\usepackage{bbm}
\usepackage{algorithm,algorithmic}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{bbm}
\usepackage[titletoc]{appendix}
\usepackage{wrapfig}
\usepackage{afterpage}
\usepackage{amssymb}

@@

\title{Semi-supervised low-rank logistic regression for
high-dimensional neuroimaging data}

@@

\author{Danilo Bzdok, Michael Eickenberg, Olivier, Grisel,
  Bertrand Thirion,
  Ga\"el Varoquaux \\\\\textbf{\textit{email:} }firstname.lastname@inria.fr}

@@

\section{Introduction}
%
Methods used in neuroimaging research can be grouped by discovering
neurobiological structure or revealing the neural correlates associated
with mental tasks.
To discover coherent distributions of activation structure across time,
independent component analysis (ICA; \cite{beckmann2005}) is often used
to decompose the BOLD (blood-oxygen level-dependent) signals into the
important modes of variation.
The ensuing spatial activation patterns are believed to represent
brain networks of
functionally interacting brain regions.
Similarly, sparse principal component analysis (SPCA; \cite{varoqu2011})
has been used to
separate brain activity signals into parsimonious network components.
Thus extracted brain networks have been shown to be
manifestations of electrophysiological oscillation frequencies \cite{hipp15}.
Their fundamental role in brain organization is
attested by continued covariation during sleep and anesthesia.
%
Network discovery is typically performed by applying ICA or SPCA on
unlabeled "resting-state" data. These capture brain dynamics
during ongoing random thought without controlled environmental input.
The biggest fraction of the BOLD signals are known
not to correlate with a particular behavior, stimulus, or experimental task.

@@

On the other hand, to investigate
the neural correlates underlying mental tasks,
the general linear model (GLM; \cite{friston94}) is the dominant approach.
The contribution of
individual brain voxels is estimated
according to a design matrix of experimental tasks.
Alternatively, psychophysiological interactions (PPI; \cite{friston97}),
elucidate the functional interactions between voxels as a function
of experimental tasks.
Dynamic causal modeling (DCM; \cite{stephan04}), in turn, quantifies directed,
task-driven influences between regions
by treating the brain as a nonlinear dynamic system with unknown
neuronal states. As a last example, always more neuroimaging studies model
experimental tasks by training classification algorithms on brain signals
\cite{poldrack09decoding}.
All these methods are applied to labeled task data that capture brain dynamics
during stimulus-guided behavior.
Two important conclusions can be drawn.
First, the mentioned supervised neuroimaging analyses operate
without exception in a single-voxel space. This ignores the fact that the BOLD
signal exhibits coherent spatial activation patterns.
Second, existing neuroimaging analyses do not acknowledge the fact that the
task-induced changes of the BOLD signal amount to less than 5\%
of baseline activity \cite{fox07}. They do thus not exploit the high similarity
of BOLD dynamics in the human brain at rest and during experimental tasks.
Indeed, very similar brain networks were observed when applying ICA
separately on rest and task data \cite{smith2009}.

@@

Both biological properties can be conjointly exploited in a 
semi-supervised (i.e., use rest and task data)
low-rank (i.e., perform network decomposition)
approach.
%
The integration of brain-network discovery in a 
supervised classification goal should identify the
neurobiological structure that allows for the best predictive models.
%
Autoencoders suggest themselves because they can emulate
variants of most nonsupervised learning algorithms,
including PCA, SPCA, and ICA.
Autoencoders
are one-layered learning models that condense the input data to
local and global representations
by improving reconstruction from them \cite{hinton06}.
%
They behave like a PCA
in case of one linear hidden layer and a squared error loss
\cite{baldi1989neural}.
This architecture yields a convex optimization objective
with unique global minimum.
Autoencoders behave like a SPCA if shrinkage terms are added for the
matrix weights in the optimization objective.
In turn, they behave like an ICA in case of a nonlinear convex
function of the first-layer activation and tied weights \cite{le2011ica}.
These authors further demonstrated that ICA, sparse autoencoders, and 
sparse coding are mathematically equivalent
under mild conditions.
In this way, autoencoders can flexibly project the neuroimaging data
onto the axes of main variation and thus
reverse-engineer properties of the data-generating
neural processes \cite{olshausen96}.

@@

In the present investigation,
an autoencoder will be fed by (unlabeled) rest data and
integrated as bottleneck
into a low-rank logistic regression fed by (labeled) task data.
Using the chain rule in back-propagation, we can then
solve the unsupervised data representation and a supervised classification
in an identical statistical learning problem.
%
From the perspective of dictionary learning, the first layer can be
viewed as a learned set of basis functions
whose linear combinations are learned
in the second layer \cite{olshausen96}.
%
Neurobiologically, this allows 
delineating a low-dimensional manifold of brain network patterns and then 
distinguishing mental tasks
by their most discriminative linear combinations.
%
Theoretically, a big reduction of the model variance is expected by
regularization using resting-state autoencoder
to put probability mass on the most neurobiologically
valid members of the model space.
%
The generalization performance should consequently be improved due to 
much reduced Vapnik-Chervonenkis dimensions of the classification estimator.
%
Taken together,
the important modes of variation in brain dynamics and
the neural correlates subserving mental operations
have mostly been studied in isolation.
We provide a principled computational framework to link these previously
unconnected domains of systems neuroscience.

\begin{wrapfigure}{r}{0.40\textwidth}
  \begin{center}
    \includegraphics[width=0.40\textwidth]{figures/figure1.png}
  \end{center}
  \caption {\textbf{Architecture}
  }
\end{wrapfigure}

@@

linear decomposition + elastic net penalty -> has characteristics of a
sparse PCA
\begin{wrapfigure}{r}{0.40\textwidth}
  \begin{center}
    \includegraphics[width=0.40\textwidth]{figures/figure3.png}
  \end{center}
  \caption {\textbf{Equilibrium between autoencoder and low-rank logistic regression}
  One learned decomposition component (out of 20) between the only-autoencoder
  (\textit{uppermost panel}) and only-logistic-regression
  (\textit{lowermost panel}) scenarios.
  The congruent structure (\textit{red, middle column})
  in the
  posterior cingulate cortex, posterior mid-cingulate cortex, and medial
  prefrontal cortex that is hence important in decompositions of the rest and
  task neuroimaging data. With increasing weight on the supervised learning,
  non-congruent structure emerges in the early (\textit{blue}) and
  lateral (\textit{red}) visual cortex
  (\textit{right column}). Matrix weights were z-scored.
  }
\end{wrapfigure}

@@

AE:
coordinate system for points on the manifold
low-dimensional bottleneck layer

@@

whereas elasticnet on the model paramters is regulization

HINTS:
In fact, the constraint by a rest-data autoencoder qualifies as a
\textit{hint} rather than regularization \cite{abu1994hints}.
Its purpose is not to prevent overfitting but to introduce
prior knowledge on known properties of the unknown target function $f$.
Rather than relying only on input-output pairs in the learning process,
we thus narrow our hypothesis set to the biologically most plausible solutions.
That is, we reduce the search space in a way that
is compatible with the expected representation of BOLD activity signals.

@@

{arg\,min}_\theta \hspace{0.2cm} {\mathcal{L}}(\theta, \lambda) = \lambda\frac{1}{N_{X_{task}}} \sum_{i=0}^{N_{X_{task}}} \log(P(Y=y^{(i)}|x^{(i)}; W,b)) \hspace{0.2cm} + \hspace{0.2cm} (1-\lambda)\frac{1}{N_{X_{rest}}}\|X_{rest}-\hat X_{rest}(W,b)\|^2

@@

As solver, we chose \textit{rmsprop},
a mini-batch version of rprop \cite{dauphin2015rmsprop}.
This procedure dictates an adaptive learning rate
for each model parameter by
scaled gradients from a running average.
Gradient normalization by RMSprop
is known to effectively exploit curvature information.
We opted for a small batch size of $100$, given the high degree of
redundancy in $X_{rest}$ and $X_{task}$.
The matrix parameters were initalized by Gaussian random values multiplied
by a gain of $0.004$. The bias parameters were initalized to be zero.
The normalization factor is computed by

\begin{eqnarray*}
v^{(t+1)} = \rho v^{(t+1)} + (1 - \rho)(\nabla f)^2
\end{eqnarray*}

where $0 < \rho < 1$ constitutes the decay rate. $\rho$ was set to
$0.9$ to deemphasize the magnitude of the gradient.
The update step is
given by

\begin{eqnarray*}
\theta^{(t+1)} = \theta^{(t)} + \alpha \frac{\nabla f(\theta^{(t)})}{\sqrt{v^{(t+1)} + \epsilon}}
\end{eqnarray*}

where $\alpha$ is the learning rate and $\epsilon$ a damping factor.
The hyper-parameter $\alpha$ was set to $0.00001$ by prior manuel
cross-validation and $\epsilon$ was set to $1^{-6}$.
%
Note that we have also experimented with other solvers
(stochastic gradient descent, adadelta, and adagrad) but found that
rmsprop converged faster and with higher generalization performance.

@@

theano/python
The analyses were performed in Python.
We used \textit{nilearn} to handle
the high-dimensional neuroimaging data and
\textit{theano} for automatic gradient
differentiation of symbolic computation graphs.
All Python scripts that generated the results are
accessible online for reproducibility and reuse
\url{http://github.com/banilo/nips2015}.

\section{Experimental Results}

@@

reduction of n gray-matter voxels to n components

@@

\section{Discussion and Conclusion}
%
There is an increasing occurrence of high-dimensional problems in the
neuroimaging domain. This calls for new statistical learning algorithms that
behave well in large-cohort settings. Ideally, they should acknowledge
and exploit existing widely-accepted neuroscientific knowledge.
In the present work, we propose such an estimator that learns
dimensionality reduction
in a neurobiologically valid and interpretable fashion.

@@

There is much uncertainty about the most pertinent representation
of neural activation information
put probability mas shwere we expeted neurobiological strucuture

canonical set of brain networks

%
\paragraph{Acknowledgment}
{\small
Data were provided the Human Connectome Project. The study was supported
by the German National Academic Foundation (D.B.).
}
